### 2017-01-22 manuscript.rmd
minor formatting

---
bibliography: manuscript.bib
csl: nature.csl
output:
  github_document: default
---

@@

Dunn, Zapata, Munro, Siebert, Hejnol

@@

A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; Grafen:1989um; Chamberlain:2012vx; Garamszegi:2014ew; Pagel:1999gc]. These have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG. Most CFG studies have abstained from phylogenetic approaches and instead heavily rely on pairwise comparisons (Fig. 1A) that do not account for the fact that evolutionary relationships explain much of the structure of variation across species. This leaves CFG studies susceptible to serious statistical errors and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

One reason that CFG has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic approaches can lead to different results in CFG. Although the value of phylogenetic approaches has been repeatedly shown in comparative analyses of other character data [@Chamberlain:2012vx; Ricklefs:1996wb], this value is apparently not well known to the FG community since the field has until recently largely focused on within-species analyses. Many initial CFG studies have also been focused on only two species at a time, where the differences between phylogenetic and pairwise approaches are not as pronounced. As the number of species considered in each CFG study grows, the shortcomings of pairwise approaches will become far more acute.

@@

![Figure Overview](Figure_overview.svg?raw=true)

@@

![Figure Expectations](figure_expectations.svg?raw=true)

@@

How changes in animal development relate to the evolution of animal diversity is a major question in evolutionary developmental biology (EvoDevo). To address this topic, Levin *et al.* [@Levin:2016bp] analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. As their title ("The mid-developmental transition and the evolution of animal body plans") indicates, they arrived at two major conclusions. First, animal development is characterized by a well-defined mid-developmental transition. Second, this transition helps explain the evolution of features observed among distantly related animals. Specifically, they concluded that animals from different phyla exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases. Closely related animals have previously been described as having an hourglass model of gene expression, where evolutionary variance in expression is greater early and late in development than at the midpoint of development [@Kalinka:2010dw; DomazetLoso:2010cn]. Levin *et al.* conclude that this contrast between distantly and closely related animals provides biological justification for the concept of phyla and may provide a long-sought operational definition of phyla. We previously described some concerns with their interpretations of these results [@Hejnol:2016hw].

Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* [@Levin:2016bp] indicate their existing data support the opposite conclusion - that the inverse hourglass is due to large differences specific to a single lineage, the ctenophore (comb jelly). Better understanding these differences will be of interest to future work, especially in light of current uncertainty regarding the phylogenetic placement of ctenophores and their unique biology [@Martindale:2015ve, Dunn:2015ij].

>>>>

---
bibliography: manuscript.bib
csl: nature.csl
output:
  github_document: default
  pdf_document:
    dev: png
---

@@

@@

A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Chamberlain:2012vx; @Garamszegi:2014ew; @Pagel:1999gc]. These have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG. Most CFG studies have abstained from phylogenetic approaches and instead heavily rely on pairwise comparisons (Fig. 1A) that do not account for the fact that evolutionary relationships explain much of the structure of variation across species. This leaves CFG studies susceptible to serious statistical errors and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

One reason that CFG has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic approaches can lead to different results in CFG. Although the value of phylogenetic approaches has been repeatedly shown in comparative analyses of other character data [@Chamberlain:2012vx; @Ricklefs:1996wb], this value is apparently not well known to the FG community since the field has until recently largely focused on within-species analyses. Many initial CFG studies have also been focused on only two species at a time, where the differences between phylogenetic and pairwise approaches are not as pronounced. As the number of species considered in each CFG study grows, the shortcomings of pairwise approaches will become far more acute.

@@

![Figure Overview](Figure_overview.png?raw=true)

@@

![Figure Expectations](figure_expectations.png?raw=true)

@@

How changes in animal development relate to the evolution of animal diversity is a major question in evolutionary developmental biology (EvoDevo). To address this topic, Levin *et al.* [@Levin:2016bp] analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. As their title ("The mid-developmental transition and the evolution of animal body plans") indicates, they arrived at two major conclusions. First, animal development is characterized by a well-defined mid-developmental transition. Second, this transition helps explain the evolution of features observed among distantly related animals. Specifically, they concluded that animals from different phyla exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases. Closely related animals have previously been described as having an hourglass model of gene expression, where evolutionary variance in expression is greater early and late in development than at the midpoint of development [@Kalinka:2010dw; @DomazetLoso:2010cn]. Levin *et al.* conclude that this contrast between distantly and closely related animals provides biological justification for the concept of phyla and may provide a long-sought operational definition of phyla. We previously described some concerns with their interpretations of these results [@Hejnol:2016hw].

Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* [@Levin:2016bp] indicate their existing data support the opposite conclusion - that the inverse hourglass is due to large differences specific to a single lineage, the ctenophore (comb jelly). Better understanding these differences will be of interest to future work, especially in light of current uncertainty regarding the phylogenetic placement of ctenophores and their unique biology [@Martindale:2015ve, @Dunn:2015ij].
